Phylogenetics and the human microbiome.

### 2013-10-02 sbreview.tex
Revision pass through.

The human microbiome is the collection of microbes that live inside and on the surface of humans.
Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information of microbiome samples.
Much of the analytical work for these sequences involves phylogenetics, at least indirectly, and methodology has developed in a somewhat different direction than for other applications of phylogenetics.
In this paper I review the field and its methods from the perspective of a phylogeneticist, as well as describing current challenges for phylogenetics coming from this type of work.

@@

The parameter regime and focus of microbiome research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what these microbial ecologists and medical researchers have done?
The answer is simple: this system is data- and question-rich.
It absolutely requires molecular methods, as microbes are now primarily identified by their molecular sequence, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sequences derived from the sample.

Although there is something of a divide between phylogeny as practiced as part of microbial ecology on one hand and that for multicellular organisms on the other, there are many parallels between the two enterprises.
Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, and species definition.
However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the scope of the former contains an almost unlimited diversity of organisms, leading to additional problems above the usual.
The species concept is even more problematic for microbes than for larger eukaryotes and there is also considerable discussion concerning how to group them into species-like units.
Organizing microbes into a sensible taxonomy is a serious challenge, especially in the absence of obvious morphological features.

Because of this high level of diversity and challenges with species definitions, microbial ecology researchers have developed their own explicitly phylogenetic techniques for doing comparing samples rather than comparing on the level of species abundances.
Although there is some overlap with previous literature, these techniques and could be used in a wider setting and may deserve broader consideration by the phylogenetics community.

The human microbiome is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
Additionally, due to more than a century of hospital laboratory work, our knowledge about human-associated microbes relatively rich.
It is also common to manipulate the human microbiome or models thereof via intervention studies and germ-free animals, and is easier to do than for many other microbial communities.

In this review I will describe phylogenetics-related research happening in microbial ecology and contrast approaches between microbial researchers and what I think of as the typical \emph{Systematic Biology} audience.
Despite an obvious oversimplification, I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
I realize that there is substantial overlap-- for instance the microbial community is very interested in unicellular fungi, and additionally many in the SB community do work on microbes-- but this terminology will be useful for concreteness.
There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees.
I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics; indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes orders of magnitude larger than typical for eukaryotic phylogenetics.

\section{The human microbiome}
The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells.
The microbiome has remarkable metabolic potential, with an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
In this section I will briefly review what is known about the human microbiome and its effect on our health.

@@

The microbiome interacts on many levels with host phenotype \citep[reviewed in][]{cho2012human}.
For example, the gut microbiome correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to children with acute metabolic syndromes in rural Africa \citep{smith2013gut}.
Considerable attention has also been given to the interaction between the gut microbiome and obesity, although the story is yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
A causal role for microbiome leading to phenotype is established for mice: an obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}, the pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
However, these promising leads have not yet been confirmed causally or in population studies of humans \citep{zhao2013gut}.
For example, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.

@@

In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perpsective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and with the human body, a broad perspective will continue to be important.

\section{Investigating the human microbiome via sequencing}
It is now possible to assay microbial communities in high throughput using sequencing.
There are two ways of doing so.
The first is to amplify a specific gene in the genome for sequencing using polymerase chain reaction (PCR).
Scientists typically pick a ``marker'' gene in that case suitable for phylogenetic analysis.
The second is to randomly shear input DNA and/or RNA and then perform sequencing directly.
We will consistently refer to the former as a \textit{survey} and the second a \textit{metagenome}, although these words have not always been consistently used in the literature.

@@

\subsection{Microbial community estimation using marker gene surveys}
Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
Their work, and the scientists who followed them, focused on the 16S ribosomal gene (henceforth simply ``16s'') as a genetic marker.
This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled surveys of almost all bacteria regardless of whether they can be cultured.

Where Woese and colleagues labored over digestion and gel electrophoresis to infer sequences, modern researchers have the luxury of high throughput sequencing.
This can be done with a high level of multiplexing, making an explicit trade-off between depth of sequencing for each specimen and the number of specimens able to be put on the sequencer at the same time.
This has led to extensive parallelization, most recently by sequencing dozens of samples at a time on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
This leads to the question of how many sequences are needed to characterize the microbial diversity of a given environment.
To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \citep[e.g.][]{caporaso2011moving}.

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
Current sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
A final cause of noise is next-generation sequencing error: this is certainly a problem for surveys and metagenomes, but is becoming less of a problem as technology improves.
I will not address it specifically except for in the inference of operational taxonomic units as described below.

\subsection{Metagenomes}
As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific region, and thus the genetic region of a read is unknown in addition to the organism it was derived from.
Because it does not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although sequencing is known to have biases.

It is possible to use metagenomic data as an expanded set of marker genes.
That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes present in a large proportion of micro-organisms that are expected to follow the same evolutionary path.
This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the diversity of gene repertoire in microbes, this core gene set may be relatively small.
Indeed, even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
The rest of the data can be taxonomically classified \citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.

Metagenomic data is often used to infer information of a metabolic rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
Such inference is beyond the scope of this paper.

@@

In this section I consider the ways in which phylogenetics has impacted human microbiome research.
What may be most interesting for this audience is the way in which phylogenetic trees are being used to actively revise taxonomy as well as being used as a structure on which to perform sample comparison.

@@

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process is somewhat opaque \citep{kreig1984bergey}.
The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to optimize the $F$-measure of precision and recall for its taxonomic assignments.
Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, the existence of multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database can be frustrating.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy on taxonomies to derive taxonomic assignments for an unknown sequence from BLAST results of an unknown sequence onto a database of sequences with those taxonomic labels.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample from a metagenome by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to classify to organisms whose genome has been sequenced.

\subsection{The role of Operational Taxonomic Units (OTUs)}
Although there continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, a substantial part of human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly at 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.
I briefly describe the mini-industry of OTU clustering techniques to contrast with the phylogenetic literature on species delimitation \citep{pons2006sequence,yang2010bayesian}.
I will use the term \emph{OTU inference} despite the fact that there is no clear definition of the OTU concept.

There are many OTU inference methods of various speeds and strategies.
Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
Different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
There are also methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values.

The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as it such errors are especially problematic there.
With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Although abundance-weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance-weighted phylogenetic diversity measures are not used in human microbiome studies.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
\citet{mccoy2013abundance} have recently shown that partially abundance-weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the traditionally-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.

@@

\newcommand{\unifracLegend}{
    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
}
\forarxiv{
\begin{wrapfigure}[23]{r}{2.5in}
  \capstart
  \vspace{-19pt}
  \begin{center}
    \includegraphics[width=2.5in]{figures/unifrac.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\unifracLegend}
  \label{fig:unifrac}
\end{wrapfigure}
}

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another via the shortest path rather than edges that connect between samples \citep{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a special case of the earth-movers distance, is special in that it can be calculated in linear time, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
Recent work \citep{chen2012associating} has shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.

The most common way to use a pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
Indeed, the separation of two communities in a principal components plot is often used as \emph{prima facie} evidence of a difference between them
\citep[e.g.][]{lozupone2007global,costello2009bacterial,yatsunenko2012human}, while the lack of such a difference is interpreted as showing that the communities are not different overall.

>>>>

The human microbiome is the collection of microbes that live inside and on the surface of humans.
Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information from microbiome samples.
Much of the analytical work for these sequences involves phylogenetics, at least indirectly, but methodology has developed in a somewhat different direction than for other applications of phylogenetics.
In this paper I review the field and its methods from the perspective of a phylogeneticist, as well as describing current challenges for phylogenetics coming from this type of work.

@@

The parameter regime and focus of microbiome research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
The answer is simple: this system is data- and question-rich.
It absolutely requires molecular methods, because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.

Although there is something of a divide between phylogeny as practiced as part of microbial ecology on one hand and that for multicellular organisms on the other, there are many parallels between the two enterprises.
Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, and species definition.
However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the scope of the former contains an almost unlimited diversity of organisms, leading to additional problems above the usual.
The species concept is even more problematic for microbes than for multicellular organisms, and hence there is also considerable discussion concerning how to group them into species-like units.
Organizing microbes into a sensible taxonomy is a serious challenge, especially in the absence of obvious morphological features.

Because of this high level of diversity and challenges with species definitions, microbial ecology researchers have developed their own explicitly phylogenetic techniques for comparing samples rather than comparing on the level of species abundances.
Although there is some overlap with previous literature, these techniques and could be used in a wider setting and may deserve broader consideration by the phylogenetics community.

The human microbiome is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
Additionally, due to more than a century of hospital laboratory work, our knowledge about human-associated microbes is relatively rich.
It is also common to manipulate the human microbiome or animal models thereof via intervention studies and germ-free animals, which is easier to do than for many other microbial communities.

In this review I will describe phylogenetics-related research happening in microbial ecology and contrast approaches between microbial researchers and what I think of as the typical \emph{Systematic Biology} audience.
Despite an obvious oversimplification, I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
I realize that there is substantial overlap--- for instance the microbial community is very interested in unicellular fungi, and additionally many in the SB community do work on microbes--- but this terminology will be useful for concreteness.
There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees.
I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics, indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes an order of magnitude larger than typical for eukaryotic phylogenetics.

\section{The human microbiome}
The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells of that individual.
The microbiome has remarkable metabolic potential, with an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
In this section I will briefly review what is known about the human microbiome and its effect on our health.

@@

The microbiome interacts on many levels with host phenotype \citep[reviewed in][]{cho2012human}.
The gut microbiome in particular correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to children with acute metabolic dysfunction in rural Africa \citep{smith2013gut}.
Considerable attention has also been given to the interaction between the gut microbiome and obesity, although the story is not yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
A causal role for the microbiome leading to obesity is established for mice: an obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}, the pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
However, these promising leads have not yet been confirmed causally or in population studies of humans \citep{zhao2013gut}.
For example, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.

@@

In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perpsective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.

\section{Investigating the human microbiome via sequencing}
It is now possible to assay microbial communities in high throughput using sequencing.
One way is to amplify a specific gene in the genome for sequencing using polymerase chain reaction (PCR).
Scientists typically pick a ``marker'' gene in that case that is meant to recapitulate the ``overall'' evolutionary history of the microbes.
Another way is to randomly shear input DNA and/or RNA and then perform sequencing directly.
We will consistently refer to the former as a \textit{survey} and the second a \textit{metagenome}, although these words have not always been consistently used in the literature.

@@

\subsection{Microbial community estimation using marker gene surveys}
Our modern knowledge of the microbial world is in a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
Their work, and the scientists who followed them, focused on the 16S ribosomal gene (henceforth simply ``16S'') as a genetic marker.
This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled surveys of almost all bacteria regardless of whether they can be cultured.

Where Woese and colleagues labored over digestion and gel electrophoresis to infer sequences, modern researchers have the luxury of high throughput sequencing.
This can be done with a high level of multiplexing, making an explicit trade-off between depth of sequencing for each specimen and the number of specimens able to be put on the sequencer at the same time.
This has led to extensive parallelization, most recently by sequencing dozens of samples at a time on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
This brings up the question of how many sequences are needed to characterize the microbial diversity of a given environment.
To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \citep[e.g.][]{caporaso2011moving}.

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for applications of marker gene sequencing to take a census of microbes.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
Current high throughput sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
I will not address it specifically except for in the inference of operational taxonomic units as described below.

\subsection{Metagenomes}
As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although multiplex sequencing is known to have biases of its own.

It is possible to use metagenomic data as an expanded set of marker genes.
That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the diversity of gene repertoire in microbes, this core gene set may be relatively small.
Indeed, even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
At least some portion of the rest of the data can be taxonomically classified \citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.

Metagenomic data is often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
Discussing these methods is beyond the scope of this paper.

@@

In this section I consider the ways in which phylogenetic methodology has impacted human microbiome research.
What may be most interesting for the \textit{Systematic Biology} audience is the way in which phylogenetic trees are being used to actively revise taxonomy as well as being used as a structure on which to perform sample comparison.

@@

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process appears opaque \citep{kreig1984bergey}.
The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to optimize the $F$-measure of precision and recall for its taxonomic assignments.
Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy on taxonomies to derive taxonomic assignments for that sequence from BLAST results of an unknown sequence onto a database of sequences with taxonomic labels.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.

\subsection{The role of Operational Taxonomic Units (OTUs)}
Although there continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, a substantial part of human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly at 97\% sequence identity, such that each OTU is a cluster of sequences that are closer to each other than that cutoff.
It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.
I briefly describe the mini-industry of OTU clustering techniques to contrast with the phylogenetic literature on species delimitation \citep{pons2006sequence,yang2010bayesian}.
I will use the term \emph{OTU inference} despite the fact that there is no clear definition of the OTU concept.

There are many OTU inference methods with various speeds and strategies.
Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
\citet{white2010alignment} show that different ways of doing this heuristic clustering can result in very different results.
\citet{navlakha2009finding} develop methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values.

The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as it such errors are especially problematic there.
With the exception of \citet{sharpton2011phylotu}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Although abundance-weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance-weighted phylogenetic diversity measures are not used in human microbiome studies.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
\citet{mccoy2013abundance} have recently shown that partially abundance-weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.

@@

\newcommand{\unifracLegend}{
    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
}
\forarxiv{
\begin{wrapfigure}[23]{r}{2.5in}
  % \capstart
  \vspace{-19pt}
  \begin{center}
    \includegraphics[width=2.5in]{figures/unifrac.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\unifracLegend}
  \label{fig:unifrac}
\end{wrapfigure}
}

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a special case of the earth-movers distance, is special in that it can be calculated in linear time, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree.
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.

The most common way to use a between-sample pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
Indeed, the separation of two communities in a principal components plot is often used as \emph{prima facie} evidence of a difference between them
\citep[e.g.][]{lozupone2007global,costello2009bacterial,yatsunenko2012human}, while the lack of such a difference is interpreted as showing that the communities are not different overall.
